\textbf{ Matching the forecast horizon with the relevant ecological processes }

### 2019-07-24 doc/manuscript.tex
Editorial changes to Discussion

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). The full model description is provided in Appendix \ref{eco-evo}. Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" (Appendix \ref{models}) predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C). We then imposed a period of warming, followed by a final period of higher stationary temperature (Fig. \ref{fig:forecast} top).

With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The temporal model accurately predicted the impact of the initial warming trend, but eventually became too pessimistic, while the spatial model did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom).

As in the community turnover example, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. The combined model initially reflected the temporal model, but then rapidly transitioned to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occurred during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The combined model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. To generate a true forecast from the combined model, we would need an independent method to predict how the model weights shift over time.

@@

\subsection*{Discussion}

Our simulations demonstrate three lessons:

\emph{Lesson 1: For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require understanding slow 
processes such as evolutionary and ecological selection as well as dispersal.} [Ethan, I am hoping you can fix and add to this paragraph, perhaps
drawing from examples in other fields?] In the short-term, time-series approaches work well because lag effects are strong: the state of the system in the next 
time step depends heavily on the current state. Furthermore, in the short-term we usually are not extrapolating far into novel conditions; the historical range of variation 
may be a good guide when there has been little time for genotype frequencies or species composition to change. In contrast, farther we forecast into the future, 
the weaker the influence of the current state, and the more likely that slow processes like selection and dispersal will alter genotype frequencies and
species composition.

This lesson has implications for the emerging field of ecological forecasting. If we only focus on iterative, near-term forecasting, we are likely to find 
that time-series approaches work well, and that consideration of slow processes like evolution, dispersal, and compositional turnover rarely improves
forecast skill. Assuming that the same models will make good longer-term forecasts would be a mistake.

We do not mean to undermine the call for an emphasis on near-term forecasting. Building capacity to make near-term forecasts is extremely important, not
only because of the many critical applications (examples?) but because it is a necessary step to building a disciplinary culture and practice of ecological 
forecasting (cite Dietze?). It is also feasible, because we have the data to validate forecasts at short time-scales. Rather, the point is that near-term forecasting
alone is not enough. We also need to build capacity to make forecasts at time-scales where slow ecological processes cannot be ignored. The challenge 
will be finding ways to validate long-term predictions.

\emph{Lesson 2. Different kinds of data reflect the operation of different processes; the choice of data for model training should depend on the time-scale of the desired forecast}.
Longitudinal data capture autocorrelation and responses of current assemblages, while data spanning spatial gradients capture the long-term outcome of 
selection and dispersal. For short-term forecasts, building models using longitudinal data is clearly the best choice. As our simulations show, models fit to 
steady-state, spatial patterns make terrible predictions in the short term. Conversely, models based on longitudinal data make poor predictions in the long-term,
while the models fit to spatial patterns can perform well. Although the temporal and spatial models we fit are phenomenological, they capture different kinds
 of underlying processes. The traditional debate about phenomenological vs. process-based models obscures the fact that, depending on the choice of the
 training data, phenomenological models can represent different kinds of processes.

 Why not simply build a process-based model? Perhaps something like the process-based models we used to generate the simulated time series in the first place? 
 That kind of model could make good predictions at both short and long time-scales. Unfortunately, this approach is just not feasible in most cases.
 The first challenge is simply determining the critical process. Take our eco-evolutionary example: should we simply assume that every annual plant species
 has warm and cold-adapted genotypes? And that temperature adaptation will be the key trait driving the response to warming, as opposed to the effects of 
 compositional turnover highlighted in our second example? Second, even if the key process is known in advance, fitting a mechanistic model presents its
 own set of challenges. The model for our eco-evolutionary example requires, in order of increasing empirical difficulty, demographic rates for the full life
 cycle, including density dependence, a quantitative mapping of genotype-by-environment interactions onto fitness, and dispersal kernels. The model for 
 our community and ecosystem example requires demography, environment-dependent intra- and interspecific competition coefficients, and dispersal kernels.
 We can think of empirical models that capture some of these features, but not all of them. Finally, even were it possible to fit such a model, the high number of 
 parameters and associated parameter error might lead to poor predictive skill (refs?).

@@

\emph{Lesson 3. A key challenge for future research is determining the rate at which slow processes first influence and ultimately dominate dynamics.}
Our simulated examples suggest the potential for a combined forecast, which weights predictions from spatial and temporal models, to 
accurately forecast both short and long-term dynamics. But we cheated: our combined ``forecasts'' were fit to data, not generated prospectively
using \emph{a priori} model weights. Finding a general solution for determining those weights poses a fascinating research problem whose 
solution would lead to important applications in ecological forecasting.

We see a role for new theory and new empirical tests. On the theoretical side, what are the parameters that determine how quickly slow processes
influence dynamics? We can play with our simulation models to provide some qualitative answers. The effects of some parameters are intuitive: in the community turnover example, increasing the fraction of dispersing individuals causes a more rapid shift in species composition and in model weights (Fig. \ref{fig:dispersal_niche_width}A). But other parameters have less straightforward effects. In the eco-evolutionary model, a potentially important parameter is the temperature tolerance of each genotype. We initially thought that increasing the temperature tolerance might accelerate shifts in genotype frequencies by increasing genotypic diversity within populations, but it had the opposite effect (Fig. \ref{fig:dispersal_niche_width}B), probably because it decreased the strength of selection, slowing the rate of change in genotype frequencies. Other factors to consider include organism lifespans and the magnitude of directional environmental change relative to interannaul variation. While simulation experiments are easy and informative, they cannot provide the generality of analytical theory. Unfortunately, deriving analytical solutions would be difficult, even for our relatively simple models.

On the empirical side, the key problem is finding or creating data sets to estimate the weighting function. As described under Lesson 1,
this requires data sets appropriate for validating long-term forecasts.
Possibilities include hindcasting with paleoecological and long-term historical data sets, use of model
systems with short-generation times, and....more simulation studies...? . With enough case studies, we might be able to infer patterns
that could guide applications in new systems.

Anything else to add?

>>>>

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). The full model description is provided in Appendix \ref{eco-evo}, and all computer code is available on Github  (\texttt{https://github.com/pbadler/space-time-forecast}). Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" (Appendix \ref{models}) predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C).

We then imposed a period of warming, followed by a final period of higher, but stationary, temperature (Fig. \ref{fig:forecast} top). With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The implications for forecasting are clear: In the initial stages of the warming trend, the temporal model fit to the original, baseline conditions should make better the best predictions about resulting population trends, but in the longer term, as evolutionary change occurs, the spatial model ultimately makes better predictions.

As in the eco-evolutionary example, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. 
The temporal model accurately predicted the impact of the initial warming trend, but eventually became far too pessimistic, while the spatial model did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom). The combined model initially reflected the temporal model, but then rapidly transitioned to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occured during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The combined model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. 
To generate a true forecast from the combined model, we would need an independent method to predict how the model weights shift over time.

@@

\subsection*{Discussion}
Ecological forecasts are typically made using models fit to either time-series data or spatial data based on the concept of space-for-time substitutions.
Our results demonstrate that forecasts based on these two approaches can lead to very different predictions for the future state of ecological systems.
Which approach provides the most accurate forecasts depends on the time-horizon, or how far into the future we are attempting to forecast.
For short-term forecasts, time-series approaches perform best, whereas longer-term forecasts are more accurately made using models from spatial data.
This stems from the different data types capturing information on the operation of different processes.
Local time-series data capture autocorrelation (including lagged responses) and responses of current assemblages to small changes in environmental conditions.
In addition, the state of the system in the next time step depends heavily on the current state (i.e., the initial condition).  
Since short-term forecasts do not typically require extrapolating into novel conditions; the historical range of variation, incorporating lags and accurate
initial conditions present in local time-series data is likely to be the most generally userful approach for forecasting.
In contrast, data spanning spatial gradients captures the long-term outcome of slower processes, like selection and dispersal. This makes models based on these
data more likely to accurately predict long-term steady state outcomes of ecological change.

These results suggest that near-term forecasts will likely be most effective when made using time-series models, a result backed by recent empirical
analyses of biodiversity forecasts at time scales from one to ten years (Harris et al. 2018). For long-term forecasts our results suggest the use
of models built on spatial data assuming. However, determining what is "near-term" and what is "long-term" in this context is an open question, and
many important forecasts in ecology will be needed at intermediate time horizons (on the scale of one to two decades) where neither approach will be
fully sufficient. Our simulation results demonstrate that there are often extended transitional period where neither approach is effective on its own,
but where a weighted combination of time-series and spatial models can produce highly accurate predictions. Unfortunately there is currently no clear
way to determine how to set the weights for the different approaches for different time horizons. Our combined ``forecasts'' used the performance of
the forecasts on simulated data to select the weights and therefore our approach cannot be used for actual forecasting. Determining how to set these
weights could be pursued using both theoretical and empirical approaches. On the theoretical side, it should be possible to explore the influence of different
parameters on how quickly slow processes (represented in spatial data) begin to influence dynamics. The effects of some parameters are intuitive:
in the community turnover example, increasing the fraction of dispersing individuals causes a more rapid shift in species composition and in model weights
(Fig. \ref{fig:dispersal_niche_width}A). Other factors to consider include organism lifespans and the magnitude of directional environmental change relative
to interannaul variation. From an empirical perspective, we can address this issue by creating data sets to estimate the weighting function. By evaluating both time-series and
spatial models for large numbers of forecasts across a range of time-horizons it should be possible to be begin to develop rules of thumb for ensembling
these two modeling approaches. This requires data sets that can support the evaluation of long-term forecasts, which could include paleoecological data (e.g.,),
long-term historical data sets (e.g., ), and the use of model systems with short-generation times (e.g., Petchy), With enough case studies, we might be able
to infer patterns that could guide applications in new systems.

Similar challenges with intermediate time horizons occur in other forecasting contexts. Weather forecasts based one type of models are very effective for
forecasting a week to ten days in advance, but then become largely uninformative. Climate models provide accurate predictions for similar variables, but
generally at decadal time-scales. Forecasting in between these two time scales has been challenging.

For short-term forecasts, building models using time-series data will often be the best choice when this type of data is available. As our simulations show,

require understanding slow 
processes such as evolutionary and ecological selection as well as dispersal.} [Ethan, I am hoping you can fix and add to this paragraph, perhaps
drawing from examples in other fields?] In contrast, the further we forecast into the future, 
the weaker the influence of the current state, and the more likely that slow processes like selection and dispersal will alter genotype frequencies and
species composition.

; the choice of data for model training should depend on the time-scale of the desired forecast.

One potential solution to these challenges is to build fully process based models of ecological systems. If it is possible to accurately characterize all
of the processes governing a system then a model based on that understanding should make accurate predictions at all time-horizons. For example, if we had
included the process-based models that we used to generate the simulated time series, these models would have been able to accurately capture
the full dynamics and forecasts for all time horizons. In concept, process based models should also be more robust for making predictions outside of historically
observed conditions (or beyond the conditions observed across spatial gradients). This is particularly important for making predictions in a future that is
predicted to be increasingly "non-analog", where combinations of environment and species interactions are operating that have never occured before (Williams et al.)
Unfortunately, this approach is not currently feasible in most cases, because we lack a detailed
knowledge of all of the complex and interacting processes influencing the dynamics of real systems...
The first challenge is simply determining the critical process. Take our eco-evolutionary example: should we simply assume that every annual plant species
has warm and cold-adapted genotypes? And that temperature adaptation will be the key trait driving the response to warming, as opposed to the effects of 
compositional turnover highlighted in our second example? Second, even if the key process is known in advance, fitting a mechanistic model presents its
own set of challenges. The model for our eco-evolutionary example requires, in order of increasing empirical difficulty, demographic rates for the full life
cycle, including density dependence, a quantitative mapping of genotype-by-environment interactions onto fitness, and dispersal kernels. The model for 
our community and ecosystem example requires demography, environment-dependent intra- and interspecific competition coefficients, and dispersal kernels.
We can think of empirical models that capture some of these features, but not all of them. Finally, even were it possible to fit such a model, the high number of 
parameters and associated parameter error might lead to poor predictive skill (refs?)... In contrast, for our simulations models based on time-series data make poor predictions in the long-term,
while the models fit to spatial patterns can perform well. Although the temporal and spatial models we fit are phenomenological, they capture different kinds
 of underlying processes. The traditional debate about phenomenological vs. process-based models obscures the fact that, depending on the choice of the
 training data, phenomenological models can represent different kinds of processes.

@@

These results have important implications for the emerging field of ecological forecasting. If we only focus on iterative, near-term forecasting, we are likely to find 
that time-series approaches work well, and that consideration of slow processes like evolution, dispersal, and compositional turnover rarely improves
forecast skill. Assuming that the same models will make good longer-term forecasts would be a mistake. We do not mean to undermine the call for an emphasis
on near-term forecasting. Building capacity to make near-term forecasts is extremely important, not
only because of the many critical applications (examples?) but because it is a necessary step to building a disciplinary culture and practice of ecological 
forecasting (cite Dietze?). It is also feasible, because we have the data to validate forecasts at short time-scales. Rather, the point is that near-term forecasting
alone is not enough. We also need to build capacity to make forecasts at time-scales where slow ecological processes cannot be ignored. The challenge 
will be finding ways to validate long-term predictions.
